Direct Entropy Measurement in a Mesoscopic Quantum System

### 2018-01-27 paper/draft.tex
Josh edits of Jan 26

Measuring the entropy of an electronic state is a powerful tool for identifying its underlying microscopic character.  Such measurements are typically based on bulk properties, such as heat capacity, that are straightforward to observe in macroscopic samples but exceedingly difficult to access in mesoscopic systems, sometimes consisting of just a few electrons. Taking advantage of a well-known Maxwell relation, we realize a mesoscopic geometry in a GaAs heterostructure in which entropy-to-charge conversion enables an entropy measurement of quantum states down to the single electron level. As a demonstration of the efficacy of this method, the entropy of the first few-electron spin states in a gate-defined GaAs quantum dot are measured. The entropy of a single spin ($k_B \ln{2}$) is measured within 8\% accuracy, as is the entropy arising at the magnetic field-driven singlet-triplet crossing for two electrons.

@@

While measurements of heat capacity are effective in probing the entropy of macroscopic 3D samples, the heat capacities of single electrons or quasiparticles are much too small to measure directly; even the 2D sheet of quasiparticles in $\nu = \frac{5}{2}$ fractional quantum Hall samples has a heat capacity that vanishes in comparison to the host crystal.  Rather than attempt to resolve such infinitesimal signals, the problem can be avoided by converting a change in entropy to changes in a more easily measured quantity, charge.  Our approach is analogous to the milestone of spin-to-charge conversion, first demonstrated in GaAs quantum dots in 2004, by which undetectably-small magnetic moments of a single spin were transformed into thepresence or absence of an electron charge, which is easily detected4 \cite{Elzerman2004, Ono2004}.

In order to accomplish this entropy to charge conversion, we make use of the Maxwell relation
%
\begin{align}
\label{eqn:max}
        \left(\frac{\partial \mu}{\partial T}\right)_{p,N} &= -\left(\frac{\partial S}{\partial N}\right)_{p,T}
\end{align}
%
that connects changes in entropy, $S$, (as well as particle number $N$ and temperature $T$) to changes in the chemical potential, $\mu$, a quantity that is simple to measure and control. The fixed pressure condition is met by working well below the Fermi temperature, $T_F \sim$\SI{5}{\kelvin}, where the pressure is simply the degeneracy pressure \cite{Landau1969}. Refs.~\onlinecite{Cooper2009} and ~\onlinecite{Ben-Shach2013} propose schemes to measure the entropy derivative, $\frac{\partial S}{\partial N}$, of the $\nu = \frac{5}{2}$ state via this Maxwell relation, detecting changes in the quasiparticle chemical potential with temperature.  Our work builds on these ideas to measure the entropy of the few-electron ground states of a quantum dot. Unlike the $\nu = \frac{5}{2}$ quasiparticles described in the theoretical work, the few-electron quantum dot states studied here have an entropy that is well-understood from the spin degree of freedom \cite{Tarucha1996, Ciorga2000, Duncan2000, Lindemann2002, Potok2003, Hofmann2016}. This work demonstrates a direct thermodyanmic measurement on a simple few-particle system; the method can be easily extended as a probe into the nature of more exotic systems ,such as topologically non-trivial Majorana and other non-Abelian states.

The mesoscopic circuit shown in Fig.~\ref{fig:fig1}a realizes an electron reservoir (central region) in thermal equilibrium with a few-electron quantum dot coupled to its right side, where tunneling between the two is measured using a charge sensing QPC capacitively coupled to the dot \cite{Staring2007, Frolov2009, Thierschmann2015}. The number of electrons on the dot can be tuned using the gate voltage, $V_G$. An electron is added to an $N$ electron dot when the chemical potential to add an electron, $\mu_{N+1}$ (proportional to $V_G$), drops below the Fermi level of the reservoir, $E_F$.  Figure 1b illustrates a such a transition, as measured by the charge sensor conductance, as a function of $V_G$. The gate voltage corresponding to midpoint of the transition, $V_0$, marks the chemical potential where the probabilties of finding $N$ or $N+1$ electrons on the dot are equal. This transition is thermally broadened by the reservoir temperature $T_{res}$. Additionally, an entropy change from the $N$ to $N+1$ ground states of the dot can shift the midpoint of the transition, $V_0$, by changing the chemical potential on the dot as in Eq.~\ref{eqn:max}.  Detecting the shift in $V_0$ with respect to temperature (${\propto}\frac{\partial \mu}{\partial T}$), amounts to a measurement of the entropy difference $dS$ between the $N$ to $N+1$ ground states ($dN=1$).  In practice, the shift in $V_0$ is measured by oscillating $T_{res}$ and monitoring changes in $g_{sens}$ with a lockin amplifier.  Consider the charge transitions in Fig.~\ref{fig:fig1}b, if the warmer transition is not only broadened but shifted ($dS\neq0$), the change in $g_{sens}$ is asymmetric as shown in Fig.~\ref{fig:fig1}c; if the warmer transition is broadened but not shifted with respect to the cooler one ($dS=0$), the lineshape is anti-symmetric.

An explanation based on detailed balance is helpful in understanding how a shift in $\mu_{N+1}$ (relative to $E_F$) relates to entropy. Tunneling between the dot and reservoir is possible when the chemical potential to add an electron, $\mu_{N+1}$, is within ${\sim}3.5 k_B T$ of Fermi level of the reservoir. Away from that window, tunneling is suppressed due to a lack of available states in the reservoir. To illustrate how the tunnel rates, $\Gamma_{in}=\Gamma_{N\rightarrow N+1}$ and $\Gamma_{out}=\Gamma_{N+1\rightarrow N}$ depend on degeneracy, consider the filling of the first electron state. The 1-electron ground state has a degeneracy of 2 due to a single, unpaired spin-$\frac{1}{2}$; electrons tunneling into this state have both spin-up and -down states available. Conversely, an electron tunneling off of the dot must tunnel into a state with its same spin, that is, only half of the reservoir states are available to it. As a result, $\Gamma_{in} = 2\Gamma_{out}$ when $\mu_{N+1}=E_F$. Extending this logic to a general case gives the same results as derived from detailed balance \cite{Gustavsson2009}, 
%
\begin{align}
	\Gamma_{in} &=  g_{N+1} \Gamma f(E_F - \mu_{N+1}) \label{eqn:gin}\\
	\Gamma_{out} &= g_{N} \Gamma [1 - f(E_F - \mu_{N+1})] \label{eqn:gout}
\end{align}
%
where $\Gamma$ is a coupling constant and $g_{N/N+1}$ are the degeneracies of the $N$ and $N+1$ electron states.. The midpoint of the transition between the $N$ and $N+1$ electon states, measured by the charge sensor, occurs when the $\Gamma_{in} = \Gamma_{out}$. It is clear from Eqs.~\ref{eqn:gin} and \ref{eqn:gout}, that in the case $g_{N}=g_{N+1}$, $\Gamma_{in} = \Gamma_{out}$ when $\mu_{N+1} = E_F$. However, if the two occupation states differ in their degeneracy, the midpoint of the transition shifts away from $\mu_{N+1}=E_F$. Thermal broadening of the Fermi functions in Eqs.~\ref{eqn:gin} and \ref{eqn:gout} means that the chemical potential at which $\Gamma_{in} = \Gamma_{out}$ depends on $T$ as well as $g_{N/N+1}$. By measuring $dV_0$ ($d_\mu$) as a function of $dT$, the ratio $g_{N+1}/g_{N}$ can be determined, which is related to $dS$ through the Boltzmann entropy. The relationship between degeneracies and tunnel rates was recently explored in a time-resolved measurement \cite{Hofmann2016}.  The thermal technique presented here extends the measurement of entropy (degeneracy) to a wider set of applications, where the tunneling process may not be convenient or even possible to observe directly.

>>>>

Measuring the entropy of an electronic state is a powerful tool for identifying its underlying microscopic character.  Such measurements are typically based on bulk properties, such as heat capacity, that are straightforward to observe in macroscopic samples but exceedingly difficult to access in mesoscopic systems that may consist of just a few electrons. Taking advantage of a well-known Maxwell relation, we realize a protocol for entropy-to-charge conversation in a gate-defined GaAs quantum dot that enables an entropy measurement of the first three quantum states in to the dot. The entropy of a single spin ($k_B \ln{2}$) is measured within 8\% accuracy, as is the entropy arising at the magnetic field-driven singlet-triplet crossing for two electrons.

@@

Entropy can be probed effectively in 3D macroscopic samples via measurements of heat capacity, but the heat capacities of single electrons or quasiparticles are much too small to measure directly; even the 2D sheet of quasiparticles in $\nu = \frac{5}{2}$ fractional quantum Hall samples has a heat capacity that vanishes in comparison to that of the host crystal.  Rather than attempt to resolve such infinitesimal signals, the problem can be avoided by converting a change in entropy to changes in charge--a quantity that is easily detected at the single particle level.  Our approach is analogous to the milestone of spin-to-charge conversion, first demonstrated in GaAs quantum dots in 2004, by which undetectably-small magnetic moments of a single spin were transformed into the presence or absence of an electron charge \cite{Elzerman2004, Ono2004}.

In order to accomplish this entropy to charge conversion, we make use of the Maxwell relation
%
\begin{align}
\label{eqn:max}
        \left(\frac{\partial \mu}{\partial T}\right)_{p,N} &= -\left(\frac{\partial S}{\partial N}\right)_{p,T}
\end{align}
%
that connects changes in entropy, $S$, (as well as particle number, $N$, and temperature, $T$) to changes in the chemical potential, $\mu$, a quantity that is simple to measure and control. The fixed pressure condition of Eq.~\ref{eqn:max} is effectively met by working well below the Fermi temperature, $T_F \sim$\SI{5}{\kelvin}, where the pressure is simply the degeneracy pressure \cite{Landau1969}. Refs.~\onlinecite{Cooper2009} and ~\onlinecite{Ben-Shach2013} propose schemes to measure the entropy derivative, $\frac{\partial S}{\partial N}$, of the $\nu = \frac{5}{2}$ state via this Maxwell relation, detecting changes in the quasiparticle chemical potential with temperature.  Our work builds on these ideas to measure the entropy of the few-electron ground states of a quantum dot. Unlike the $\nu = \frac{5}{2}$ quasiparticles described in the theoretical work, the few-electron quantum dot states studied here have an entropy that is well-understood from the spin degree of freedom \cite{Tarucha1996, Ciorga2000, Duncan2000, Lindemann2002, Potok2003, Hofmann2016}. Our experiment demonstrates a direct entropy measurement on a few-particle system; in the future, the method can be easily extended as a probe into the nature of more exotic systems, such as topologically non-trivial Majorana and other non-Abelian states.

The mesoscopic circuit shown in Fig.~\ref{fig:fig1}a realizes an electron reservoir (central region) in thermal equilibrium with a few-electron quantum dot coupled to its right side, whose occupation is measured using an adjacent charge sensing quantum point contact\cite{Staring2007, Frolov2009, Thierschmann2015}. The number of electrons on the dot can be tuned using the gate voltage, $V_G$, to control the chemical potential.  An electron is added to an $N$ electron dot when the chemical potential to add an electron, $\mu_{N+1}$, drops below the Fermi level of the reservoir, $E_F$.  Figure 1b illustrates a such a transition, showing the expected step in charge sensor conductance as a function of $V_G$ thermally broadened by the reservoir temperature $T_{res}$. The gate voltage corresponding to midpoint of the transition, $V_0$, marks the chemical potential where the probabilities of finding $N$ or $N+1$ electrons on the dot are equal.

Any entropy change from the $N$ to $N+1$ ground states of the dot should shift the midpoint of the transition, $V_0$, by changing the chemical potential on the dot as in Eq.~\ref{eqn:max}.  Detecting the shift in $V_0$ with respect to temperature (${\propto}\frac{\partial \mu}{\partial T}$), amounts to a measurement of the entropy difference $dS$ between the $N$ to $N+1$ ground states ($dN=1$).  In practice, the shift in $V_0$ with temperature is measured by oscillating $T_{res}$ and monitoring any resulting changes in $g_{sens}$ with a lockin amplifier.  If the transition is broadened but not shifted due to heating ($dS=0$), an antisymmetric lineshape is expected from this lockin measurement.  If the transition is not only broadened but also shifted ($dS\neq0$), the change in $g_{sens}$ due to heating will asymmetric, as shown in Fig.~\ref{fig:fig1}c, thus offering a clearly distinguishable signature of entropy for this single charge measurement.

An explanation based on detailed balance is helpful in understanding how the shift in $\mu_{N+1}$ with respect to $E_F$ upon heating relates to entropy.   In general, electrons tunnel back and forth between the dot and reservoir as long as $\mu_{N+1}$ is within thermal broadening of Fermi level, that is, as long as there are available states in both the dot and the reservoir.  The tunnel rates into and out of the dot, $\Gamma_{in}=\Gamma_{N\rightarrow N+1}$ and $\Gamma_{out}=\Gamma_{N+1\rightarrow N}$, depend on degeneracy of the $N$ and $N+1$ levels.  Consider the filling of the first electron state at zero magnetic field, a state with a single, unpaired spin-$\frac{1}{2}$ and therefore a degeneracy of 2. An electron tunneling into this state has both spin-up and -down states available.  Tunneling off, the electron (whether spin up or down) must tunnel into reservoir states with the spin, that is, only half of the reservoir states are available to it. As a result, $\Gamma_{in} = 2\Gamma_{out}$ when $\mu_{N+1}=E_F$. Extending this logic to a general case gives the same results as derived from detailed balance \cite{Gustavsson2009}, 
%
\begin{align}
	\Gamma_{in} &=  g_{N+1} \Gamma f(E_F - \mu_{N+1}) \label{eqn:gin}\\
	\Gamma_{out} &= g_{N} \Gamma [1 - f(E_F - \mu_{N+1})] \label{eqn:gout}
\end{align}
%
%MAKE THIS ONE EQUATION

where $\Gamma$ is a coupling constant, $f$ is the Fermi function, and $g_{N/N+1}$ are the degeneracies of the $N$ and $N+1$ electron states.

The midpoint of the transition between the $N$ and $N+1$ electron states, measured by the charge sensor, occurs when the dot is occupied half the time, that is, when $\Gamma_{in} = \Gamma_{out}$. From Eqs.~\ref{eqn:ginout}, $\Gamma_{in} = \Gamma_{out}$ occurs when $\mu_{N+1} = E_F$ as long as $g_{N}=g_{N+1}$, but when $g_{N}<>g_{N+1}$ the midpoint of the transition shifts away from $\mu_{N+1}=E_F$ by an amount that depends on the broadening of the Fermi function and therefore on $T$.   Thus, a measurement of how the midpoint of the transition (V_0) shifts with T allows a determination of the ratio $g_{N+1}/g_{N}$, which itself is related to $dS$ through the Boltzmann entropy formula, $S=k_B ln(g)$.  The relationship between degeneracies and tunnel rates was recently explored in a time-resolved measurement \cite{Hofmann2016}.  The thermal technique presented here extends the measurement of entropy (degeneracy) to a wider set of applications, where the tunneling process may not be convenient or even possible to observe directly.
